---
title: "HW2"
author: "D.Smutin"
date: "2024-04-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(base.dir = "~/bioinformatics/ITMO/statistics/multivar/HW3")

library(tidyverse)
library(plotly)
library(vegan)
library(readxl)
set.seed(1)
```

# Homework 3 by Smutin Daniil

## Prepare data
```{r}
df <- read_xlsx("Sleepy lizard.xlsx")

hem <- df %>% 
  dplyr::select("Tot_WBC", "Het_ABS", "Lym_ABS", 
                "H:L Ratio", "Mon_ABS", "OthG_ABS")

env <- df[,4:7]

env$Treatment <- c("normal", "modified")[env$Treatment %>% match(1:2)]

env$Habitat <- c("normal", "swale plantings", "fields with crops", "under fallow")[env$Habitat %>% match(1:4)]

env$`Landscape ID` <- c("LS1", "LS2", "LS3"
)[env$`Landscape ID` %>% match(unique(env$`Landscape ID`))]
```

## Normalizing data
```{r}
hem %>% 
  as.data.frame %>% 
  pivot_longer(cols = 1:6) %>% 
  ggplot(aes(name, value)) +
  geom_boxplot(outliers = F) +
  geom_jitter(size = .1) +
  theme_minimal()
```

OK, standardize data and center it
```{r}
center <- function(x) (x - mean(x)) #/ sd(x)
log10_1 <- function(x) log10(x+1)

hem <- hem %>% 
  apply(2, log10_1) %>% 
  #apply(2, scale) %>% 
  apply(1, center) %>% 
  apply(1, center) %>% 
  as.data.frame()

hem %>% 
  pivot_longer(cols = 1:6) %>% 
  ggplot(aes(y = value)) +
  geom_boxplot(outliers = F, aes(x = 0), width = .5) +
  geom_density() +
  facet_grid(~name, scales = "free") +
  theme_minimal()
```

Not totally normal distributions, but near them

## Removing known batch using LBSI
```{r}
batch <- df$LBSI %>% center

## measuring its influence
model_batch <- glm(
  batch ~
    hem$Tot_WBC + hem$Het_ABS + hem$Lym_ABS + 
    hem$`H:L Ratio` + hem$Mon_ABS + hem$OthG_ABS
)

plot(model_batch)
```

14, 81 and 85 ID is marked as outliers. Check it later

## Viz
Full:
```{r}
hem %>% 
  cbind(env) %>% 
  cbind(batch) %>% 
  pivot_longer(cols = 1:6) %>% 
  mutate(value = value) %>% 
  ggplot(mapping = aes("", value)) +
  geom_boxplot(aes(fill = Treatment), outliers = F) +
  geom_jitter(aes(color = batch), size = .2, alpha = .5) +
  theme_minimal() +
  facet_wrap(~name, scales = "free") +
  scale_color_viridis_c()
```

3 out of 6 categories seems to differ. In contrast, without performing double-centralization, 3-4 categories varied.

Is batch equal?
```{r}
env %>% 
  cbind(batch) %>% 
  ggplot(mapping = aes(Treatment, batch)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  theme_minimal()
```

OK, the batch is nearly equal, but can be reduced. I choose MANOVA approach for predict.
First of all, let's remove outliers

## PCoA
```{r}
df_pcoa <- hem %>% 
  prcomp()
plot(df_pcoa)
```
OK, not bad distribution

```{r}
gg <- df_pcoa[["x"]] %>% 
  cbind(env) %>% 
  mutate (LID = as.character(`Landscape ID`)) %>% 
  mutate(HBT = as.character(Habitat)) %>% 
  ggplot(aes(PC1, PC2, color = LID, shape = HBT, text = paste0("ID: ", 1:122))) +
  facet_grid(~Treatment) +
  geom_point() +
  theme_minimal()

ggplotly(gg)
```

Without logarithmization (with usage of only the centering approach), the variance of two groups become unequal

```{r, include=F}
#filtered <- c(13,21)

dfF <- hem %>% 
  cbind(env) %>% 
  cbind(batch) %>% 
  mutate(Treatment <- as.factor(Treatment))

#dfF <- dfF[-filtered,]

df_pcoa <- dfF[,1:6] %>% 
  prcomp()
plot(df_pcoa)
```

```{r}
gg <- df_pcoa[["x"]] %>% 
  cbind(env) %>% 
  cbind(batch) %>% 
  mutate (LID = as.character(`Landscape ID`)) %>% 
  mutate(HBT = as.character(Habitat)) %>% 
  ggplot(aes(PC1, PC2, color = batch, shape = Treatment, text = paste0("ID: ", 1:122))) +
  #facet_grid(~Treatment) +
  scale_color_viridis_c(name = "LBSI") +
  geom_point(size = 1, alpha = .7) +
  theme_minimal()

ggplotly(gg)
```

OK. Seems like lizards from unmodified areas have a bigger spread without row-wise standartization. With this procedure, nothing is related to their groupping. Also, there is ~no groupping or gradient according to LBSI

### Group variance
```{r}
df_PCO <-dfF[,1:6] %>% 
  vegdist(method  = "euclidean") %>% 
  betadisper(dfF$Treatment)
plot(df_PCO)
```


```{r}
anova(df_PCO)
```

Groups residuals are equal now (with scaling with different datawizard package functions it was not true)

### perMANOVA
```{r}
df_adonis <- adonis2(hem ~ dfF$Treatment*batch, method = "euclidean")
df_adonis
```

The difference is significant. But what what exactly is the reason of the differences?

### GLM
```{r}
dfF$target <- as.factor(dfF$Treatment)
df_glm <- glm(data = dfF,
              formula = target ~
                (Tot_WBC + Het_ABS + Lym_ABS + 
                `H:L Ratio` + Mon_ABS + OthG_ABS)*batch,
              family = "binomial")

summary(df_glm)
```

IDK why is it broken on OthG var

## Habitats
### Group variance
```{r}
dfH <- dfF %>% subset(Treatment == 'modified')

df_PCO <- dfH[,1:6] %>% 
  vegdist(method  = "euclidean") %>% 
  betadisper(dfH$Habitat)
plot(df_PCO)
```


```{r}
anova(df_PCO)
```

Groups residuals are equal now (with scaling with different datawizard package functions it was not true)

### perMANOVA: habitats
```{r}
df_adonis <- adonis2(dfH[,1:6] ~ dfH$Habitat*dfH$batch, method = "euclidean")
df_adonis
```

The difference is not significant across different habitats.

### perMANOVA: connectivity
```{r}
df_adonis <- adonis2(dfH[,1:6] ~ dfH$Connectivity*dfH$batch, method = "euclidean")
df_adonis
```

The difference is also not significant

### perMANOVA: connectivity and habitat
```{r}
df_adonis <- adonis2(dfH[,1:6] ~ dfH$Connectivity*dfH$Habitat*dfH$batch, method = "euclidean")
df_adonis
```

Also, no visible relations. It might be better if I somehow balance variables, but not in now.

### GLM?
```{r}
dfH$target <- as.factor(paste(dfH$Habitat, dfH$Connectivity))
df_glm <- glm(data = dfH,
              formula = target ~
                (Tot_WBC + Het_ABS + Lym_ABS + 
                `H:L Ratio` + Mon_ABS + OthG_ABS),
              family = "binomial")

summary(df_glm)
```

```{r}
plot(df_glm)
```

Several outliers could be removed to improve both GLM and perMANOVA
```{r}
dfH <- dfH[!(rownames(dfH) %in% c(70, 91, 96, 119)),]

df_glm <- glm(data = dfH,
              formula = target ~
                (Tot_WBC + Het_ABS + Lym_ABS + 
                `H:L Ratio` + Mon_ABS + OthG_ABS),
              family = "binomial")

plot(df_glm)
```

```{r}
df_adonis <- adonis2(dfH[,1:6] ~ dfH$Connectivity*dfH$Habitat*dfH$batch, method = "euclidean")
df_adonis
```

Still, no connection using AOV, so GLM might have a noisy features for the prediction.

# Answers
## Task 1
a) Yes
b) No
c) No
## Task 2
d) No